Neural Networks Loss Landscape Convergence in Hessian Low-Dimensional Space

### 2025-04-30 paper/main.tex
Universal commits presents...

\begin{enumerate}
  \item Constructing a Hessian‑based projection that captures the principal curvature directions.
  \item Applying Monte Carlo and other sampling methods to identify when the loss landscape stabilizes as dataset size grows.
  \item Visualizing the projected loss surface around local minima to illustrate these changes.
  \item Deriving theoretical criteria --- using results from The Matrix Cookbook \cite{petersen2012matrix} --- to predict when additional
        samples have minimal impact.
\end{enumerate}

@@

\textbf{Organization.} The rest of the paper is structured as follows. Section~2 reviews related work. Section~3 introduces notation
and preliminaries. Section~4 derives theoretical bounds on Hessian spectra and the loss‐difference norm. Section~5 presents our
empirical studies. We discuss the implications in Section~6 and conclude in Section~7. Additional experiments and proofs are
provided in Appendix~A.

@@

We consider a $S$-class classification problem. Let $\bx \in \bX$ be an input and $y \in \mathcal Y = \{1, \dots, S\}$ its label. A
neural network with parameters $\bw \in \bOmega \subset \mathbb R^P$ defines a mapping $f_{\bw} : \bX \to \mathbb R^S$. Given a dataset
$$
  \mathcal D =
  \{(\bx_i, y_i)\}_{i=1}^N
$$
of $m$ i.i.d. samples, and a per-sample loss $\ell \bigl( f_{\bw}(\bx_i), y_i \bigr)$ (e.g. cross–entropy), we define
$$
  \ell_i(\bw) =
  \ell \bigl( f_{\bw}(\bx_i) , y_i \bigr)\,.
$$

@@

\begin{definition}
  Fix a tolerance $\Delta > 0$. The \emph{sufficient sample size} is
  $$
    K =
    \inf \{ k \mid \forall\, k' \ge k : \; \Delta_{k'} < \Delta\}.
  $$
  When $k \ge K$, adding further samples changes the loss landscape by less than $\Delta$.
\end{definition}

@@

We make the following standard assumptions to simplify our analysis around a local minimizer~$\bw^*$.

@@

\section{Method: Projection onto Dominant Eigen-Directions and Loss Landscape Approximation}

Working with the full Hessian $\mathbf{H}_k(\bw)$ of a neural network is computationally expensive due to the high dimensionality
of the parameter space $\bOmega$. To reduce this cost, we project the parameters onto a lower‑dimensional subspace spanned by the
$d$ dominant eigenvectors of the Hessian at a local minimum $\bw^*$.

The justification for this projection is that, empirically, the Hessian has low effective rank near a minima \cite{sagun2018empirical}:
only a few eigenvalues are significantly above zero, while the rest are negligible.  Therefore, directions corresponding to small
eigenvalues contribute little to loss variation.  Retaining only the top $d$ eigenvectors yields a compact yet informative representation
of the most important curvature directions.

For illustration, Figure~\ref{fig:evgen} plots the top few eigenvalues of $\bH(\bw^*)$ in descending order. We compute these dominant
eigenvalues using the iterative power‐method described at \cite{hessian-eigenthings}.

@@

Concretely, let
$$
  \bP =
  \bigl[ \mathbf{e}_1, \dots, \mathbf{e}_d \bigr]
$$
be the matrix whose columns are the top $d$ eigenvectors.  We write any parameter vector as
$$
  \bw =
  \bw^* + \bP\, \btheta, \quad
  \btheta \in \mathbb R^d.
$$
Substituting into a second‑order Taylor expansion of $\cL_k$ around $\bw^*$ gives
$$
  \cL_k\bigl(\bw^* + \bP\,\btheta\bigr) \approx
  \cL_k(\bw^*) + \tfrac12\, \btheta^{\T}\! \bigl( \bP^{\T}\, \bH_k(\bw^*)\, \bP\bigr)\, \btheta =
  \cL_k(\bw^*) + \tfrac12\, \btheta^{\T} \bLambda_k\, \btheta,
$$
where
$$
  \bLambda_k =
  \bP^{\T}\, \bH_k(\bw^*)\, \bP =
  \operatorname{diag} \bigl( \lambda_k^{(1)}, \dots, \lambda_k^{(d)} \bigr).
$$
An identical expansion holds for $\cL_{k+1}$, yielding $\bLambda_{k+1}$ from the Hessian on $k+1$ samples.

\begin{theorem}[Approximate $\Delta_k$ via Eigenvalues]
  Under the assumption that $\btheta \sim \mathcal N(0, \sigma^2 \mathbf I_d)$, the change in loss can be estimated by
  $$
    \Delta_k =
    \mathbb E \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2 \approx
    \frac{\sigma^4}{4}\, \Biggl( 2 \sum_{i=1}^d (\lambda_{k+1}^{(i)} - \lambda_k^{(i)})^2
    + \Bigl( \sum_{i=1}^d (\lambda_{k+1}^{(i)} - \lambda_k^{(i)}) \Bigr)^2 \Biggr).
  $$
\end{theorem}

@@

This formula uses only the top $d$ eigenvalues, so it is efficient to compute even for very large models.
In practice, it provides an upper bound on the empirical $\Delta_k$ and thus a reliable criterion for determining when the
dataset size is sufficient.

\section{Experiments}

@@

\paragraph{Experimental Procedure.}
\begin{enumerate}
  \item \textbf{Random‐direction visualization:} As an initial sanity check, we project the parameters onto two random directions and
        plot the resulting 3D loss landscapes and their squared differences.
  \item \textbf{Hessian‐based analysis:} We then repeat the visualization and compute $\Delta_k$ by projecting onto the top two
        (and up to $d = 10$) eigenvectors of the Hessian.  This directly tests our eigenvalue‐based approximation and the convergence behavior
        predicted by theory.
\end{enumerate}

All code and detailed configuration files are available in our GitHub repository.

\subsection{Preliminary Experiment: Projection onto Random Directions}
\subsubsection{Visualizing the Loss Landscape}

@@

\begin{enumerate}
  \item \textbf{Grid Sampling.}
        We form a uniform grid of $(\alpha, \beta)$ pairs in $[-1,1]\times[-1,1]$ with a fixed step size.
  \item \textbf{Loss Computation.}
        For each $(\alpha,\beta)$, we set $\bw=\bw(\alpha,\beta)$, keep the base weights $\bw^*$ fixed, and compute both
        $\cL_k(\bw)$ and $\cL_{k+1}(\bw)$.
  \item \textbf{Aggregated Metrics.}
        \begin{itemize}
          \item The surface $\cL_k(\alpha,\beta)$ visualizes how the mean loss changes over the grid.
          \item The surface $\bigl( \cL_{k+1} (\alpha, \beta) - \cL_k (\alpha, \beta) \bigr)^2$ highlights where adding one sample
                causes the largest change.
        \end{itemize}
  \item \textbf{3D Visualization.}
        We render two side‐by‐side 3D plots:
        \begin{itemize}
          \item \emph{Left:} $\cL_k (\alpha, \beta)$ vs. $(\alpha, \beta)$.
          \item \emph{Right:} $\bigl( \cL_{k+1} (\alpha, \beta) - \cL_k (\alpha, \beta) \bigr)^2$ vs. $(\alpha, \beta)$.
        \end{itemize}
\end{enumerate}

@@

\paragraph{Loss Landscape for the Small Dataset}
In Figure~\ref{fig:loss_random_small}, the loss surface is highly irregular, with sharp peaks and deep valleys.
This irregularity indicates that with very few training samples, the model is extremely sensitive to small parameter perturbations.
The squared difference surface also exhibits large spikes, confirming that adding a single data point can cause substantial local changes
in the loss.

\paragraph{Loss Landscape for the Full Dataset}
In contrast, Figure~\ref{fig:loss_random_big} shows a much smoother loss surface with gentle slopes. The overall loss
values are lower --- reflecting the averaging effect of more samples --- and the squared difference surface is correspondingly much smaller.
This smoothing demonstrates that as the dataset grows, the local optimum stabilizes, and additional samples produce only minor
adjustments to the loss landscape. Hence, beyond a certain dataset size, further data have negligible impact on the shape of the loss
surface.

@@

To study how the loss landscape converges as the dataset grows, we estimate and plot the function
$$
  \Delta_k =
  \mathbb{E} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2
$$
via a Monte Carlo approximation using Gaussian perturbations in parameter space. The steps are:

\begin{enumerate}
  \item \textbf{Sample directions.}
        Starting from a fixed trained model with parameters $\bw^*$, draw $T$ independent random vectors
        $$
          \btheta^{(t)} \sim \mathcal{N} (\mathbf 0, \sigma^2 \mathbf I_d), \quad
          t = 1, \dots, T.
        $$

  \item \textbf{Perturb parameters.}
        For each sample, form a perturbed parameter set
        $$
          \bw^{(t)} =
          \bw^* + \btheta^{(t)}.
        $$

  \item \textbf{Evaluate loss differences.}
        Compute
        $$
          \cL_k \bigl( \bw^{(t)} \bigr) \quad
          \text{and} \quad
          \cL_{k+1} \bigl( \bw^{(t)} \bigr),
        $$
        then calculate the squared difference
        $$
          d_t =
          \bigl[ \cL_{k+1} (\bw^{(t)}) - \cL_k (\bw^{(t)}) \bigr]^2.
        $$

  \item \textbf{Compute Monte Carlo estimate.}
        Approximate
        $$
          \Delta_k \approx
          \frac{1}{T} \sum_{t=1}^T d_t.
        $$
        Increasing $K$ reduces the variance of this estimator.

  \item \textbf{Plot results.}
        We plot $\Delta_k$ versus $k$ to observe its decay. We also plot the scaled quantity
        $$
          \Delta_k \cdot k^2
        $$
        alongside to highlight an $\mathcal O(1 / k)$ convergence rate.
\end{enumerate}

Figures~\ref{fig:delta_random_sigma} and \ref{fig:delta_random_dim} present the behavior of the function
$$
  \Delta_k =
  \mathbb{E} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2
$$
under different Monte Carlo sampling settings and projection subspace sizes. Each figure isolates one variable:

\begin{itemize}
  \item Figure~\ref{fig:delta_random_sigma}: varying the Gaussian variance $\sigma$.
  \item Figure~\ref{fig:delta_random_dim}: varying the subspace dimension $d$.
\end{itemize}

\begin{figure}[!htbp]
  \hspace*{-2.6cm}
  \subfloat{\includegraphics[width=1.35\textwidth]{img/delta_random_sigma_2_64.pdf}}
  \caption{$\Delta_k$ (left) and $\Delta_k \cdot k^2$ (right), varying the Gaussian variance with fixed $d = 2$.}
  \label{fig:delta_random_sigma}
\end{figure}

\begin{figure}[!htbp]
  \hspace*{-2.6cm}
  \subfloat{\includegraphics[width=1.35\textwidth]{img/delta_random_dim_1_64.pdf}}
  \caption{$\Delta_k$ (left) and $\Delta_k \cdot k^2$ (right), varying the subspace dimension with fixed $\sigma = 1$.}
  \label{fig:delta_random_dim}
\end{figure}

\paragraph{Comparing Different Variances \texorpdfstring{$\sigma$}{sigma}}
Figure~\ref{fig:delta_random_sigma} shows that varying the Gaussian variance $\sigma$ produces no substantial change in the shape or
magnitude of the $\Delta_k$ curves. This insensitivity likely stems from using random projection directions: since they do not align
with the dominant curvature axes of the Hessian, scaling those directions has little impact on the estimated loss differences.

\paragraph{Comparing Different Subspace Dimensions}
Figure~\ref{fig:delta_random_dim} shows that, when using random directions, increasing the projection dimension $d$ does not
significantly alter the overall shape of $\Delta_k$. This suggests that only a few directions dominate the variation --- motivating our
focus on Hessian‐based projections in the main experiment.

In all cases, the right‐hand plots (showing $\Delta_k \cdot k^2$) exhibit a roughly $\mathcal O(1 / k)$ decay, confirming that the
incremental effect of adding one more sample diminishes proportionally to $1 / k$ as the dataset grows.

@@

\subsection{Moving Forward: Projection onto Dominant Hessian Eigenvectors}

\paragraph{Projection Directions.}
In the main experiment, we replace random directions with the $d$ eigenvectors of the Hessian at the local minimizer $\bw^*$ that
correspond to its largest eigenvalues.  Prior work \cite{sagun2018empirical} shows these top modes capture nearly all of the
non‑negligible curvature, and Figure~\ref{fig:evgen} confirms a rapid eigenvalue decay. Fixing $d=10$ balances computational cost and
fidelity, and our tests on Fashion‑MNIST and CIFAR demonstrate that this choice suffices even for complex architectures.

@@

Figures~\ref{fig:loss_eigen_small} and \ref{fig:loss_eigen_big} display 3D plots of $\cL_k(\btheta)$ (left) and 
$\bigl[ \cL_{k+1}(\btheta) - \cL_k(\btheta) \bigr]^2$ (right), where $\btheta$ varies over the span of the top two Hessian eigenvectors.
 The procedure mirrors the random‑direction case but uses these maximally informative axes.

@@

\paragraph{Loss Landscape for the Small Dataset}
In Figure~\ref{fig:loss_eigen_small}, the loss surface projected onto the top two Hessian eigenvectors is
noticeably more structured and less erratic than in the random‐direction case. This confirms that these eigenvectors capture the
most influential directions of loss variation. The squared‐difference plot also shows larger peaks, indicating that the model is
highly sensitive to new data along these key axes when $k$ is small.

\paragraph{Loss Landscape for the Full Dataset}
By contrast, Figure~\ref{fig:loss_eigen_big} shows a much smoother loss surface and a greatly reduced squared‐difference.
As the dataset grows, the loss changes less along the principal curvature directions, and the model’s local behavior stabilizes.
Notably, even at large $k$, the squared difference remains higher than in the random‐direction experiment, providing a practical upper
bound on landscape fluctuations. These findings reinforce that projecting onto the top eigenvectors effectively reveals the convergence
behavior of the loss landscape and helps intuitively understand the existence of the point beyond which additional data have minimal impact.

@@

\begin{figure}[!htbp]
  \hspace*{-2.6cm}
  \subfloat{\includegraphics[width=1.35\textwidth]{img/delta_eigen_1_10_64.pdf}}
  \caption{$\Delta_k$ (left) and $\Delta_k \cdot k^2$ (right)}
  \label{fig:delta_eigen}
\end{figure}

Figure~\ref{fig:delta_eigen} presents the Monte Carlo estimate of $\Delta_k$ in the subspace spanned by the top $d = 10$
Hessian eigenvectors. Compared to the random‐direction experiment, $\Delta_k$ here has a higher value --- reflecting the
large curvature captured by these eigenvectors. The right‐hand plot of $\Delta_k\cdot k^2$
confirms an approximate $O(1 / k)$ convergence rate. These observations show that projecting onto the leading eigen-directions not only
highlights the greatest loss sensitivity but also yields a reliable upper bound on how quickly the landscape stabilizes as more data are
added.

\subsection{Main Experiment: Theoretical vs. Empirical \texorpdfstring{$\Delta_k$}{Delta k}}

@@

We approximate $\Delta_k$ by
$$
  \Delta_k \approx
  \frac{\sigma^4}{4}\, \Biggl( 2 \sum_{i=1}^d (\lambda_{k+1}^{(i)} - \lambda_k^{(i)})^2
  + \Bigl( \sum_{i=1}^d (\lambda_{k+1}^{(i)} - \lambda_k^{(i)}) \Bigr)^2 \Biggr).
$$
where $\{\lambda_k^{(i)}\}_{i=1}^d$ and $\{\lambda_{k+1}^{(i)}\}_{i=1}^d$ are the top $d$ Hessian eigenvalues at sample sizes $k$ and $k + 1$,
and $\sigma$ is the perturbation variance.

\begin{enumerate}
  \item \textbf{Model \& subspace size.} Train a network to obtain $\bw^*$. Choose $d$ based on the rapid eigenvalue decay in
        Figure~\ref{fig:evgen}.
  \item \textbf{Eigenvalue computation.} For each $k$ in the range of interest (typically the largest $K$ sizes), compute the top
        $d$ eigenvalues via the power method.
  \item \textbf{Theoretical $\Delta_k$.} Plug these eigenvalues into the formula above to obtain a sequence of theoretical estimates.
  \item \textbf{Empirical $\Delta_k$.} Independently estimate $\Delta_k$ by Monte Carlo sampling of perturbations along the same
        top‐$d$ eigenvectors as it was in the previous experiment.
  \item \textbf{Comparison.} Plot both theoretical and empirical $\Delta_k$ on the same axes to assess how tightly the theory bounds
        the observed values.
\end{enumerate}

\begin{figure}[!htbp]
  \subfloat{\includegraphics[width=1\textwidth]{img/delta_border_1_10_64.pdf}}
  \caption{Theoretical and Empirical $\Delta_k$}
  \label{fig:delta_eigen_comparison}
\end{figure}

Figure~\ref{fig:delta_eigen_comparison} shows the theoretical estimate of $\Delta_k$ (dashed line) alongside the empirical Monte Carlo
measurements (solid line).  Across all values of $k$, the theory curve remains above the empirical one, demonstrating that our
eigenvalue‐based formula provides a valid upper bound for the true loss‐landscape changes.

>>>>

\begin{enumerate}
  \item Constructing a Hessian‑based projection that captures the principal curvature directions.
  \item Applying Monte Carlo method to identify when the loss landscape stabilizes as dataset size grows.
  \item Visualizing the projected loss surface around local minima to illustrate these changes.
  \item Deriving theoretical criteria --- using results from The Matrix Cookbook \cite{petersen2012matrix} --- to predict when additional
        samples have minimal impact.
\end{enumerate}

@@

\textbf{Organization.} The rest of the paper is structured as follows. Section 2 reviews related work. Section 3 introduces notation
and preliminaries. Section 4 derives theoretical bounds on Hessian spectra and the loss‐difference norm. Section 5 presents our
empirical studies. We discuss the implications in Section 6 and conclude in Section 7. Additional experiments and proofs are
provided in Appendix A.

@@

We consider a $S$-class classification problem. Let $\bx \in \bX$ be an input and $y \in \mathcal Y = \{1, \dots, S\}$ its label. A
neural network with parameters $\bw \in \bOmega \subset \mathbb R^M$ defines a mapping $f_{\bw} : \bX \to \mathbb R^S$. Given a dataset
$$
  \mathcal D =
  \{(\bx_i, y_i)\}_{i=1}^N
$$
of $N$ i.i.d. samples, and a per-sample loss $\ell \bigl( f_{\bw}(\bx_i), y_i \bigr)$ (e.g. cross–entropy), we define
$$
  \ell_i(\bw) =
  \ell \bigl( f_{\bw}(\bx_i) , y_i \bigr)\,.
$$

@@

\begin{definition}
  Fix a tolerance $\Delta > 0$. The \emph{$\Delta$-sufficient sample size} is
  $$
    K =
    \inf \{ k \mid \forall\, k' \ge k : \; \Delta_{k'} < \Delta\}.
  $$
  When $k \ge K$, adding further samples changes the loss landscape by less than $\Delta$.
\end{definition}

@@

We make the following standard assumptions to simplify our analysis around a local minimizer $\bw^*$.

@@

\section{Method: Projection onto Dominant Eigen-Directions and Loss Landscape Approximation}\label{sec:meth}

Working with the full Hessian $\mathbf{H}_k(\bw)$ of a neural network is computationally expensive due to the high dimensionality
of the parameter space $\bOmega$. To reduce this cost, we project the parameters onto a lower‑dimensional subspace spanned by the
$D$ dominant eigenvectors of the Hessian at a local minimum $\bw^*$.

The justification for this projection is that, empirically, the Hessian has low effective rank near a minima \cite{sagun2018empirical}:
only a few eigenvalues are significantly above zero, while the rest are negligible. Therefore, directions corresponding to small
eigenvalues contribute little to loss variation. Retaining only the top $D$ eigenvectors yields a compact yet informative representation
of the most important curvature directions.

For illustration, Figure \ref{fig:evgen} plots the top few eigenvalues of $\bH(\bw^*)$ in descending order. We compute these dominant
eigenvalues using the iterative power‐method described at \cite{hessian-eigenthings}.

@@

Concretely, let
$$
  \bP =
  \bigl[ \mathbf{e}_1, \dots, \mathbf{e}_D \bigr]
$$
be the matrix whose columns are the top $D$ eigenvectors. We write any parameter vector as
$$
  \bw =
  \bw^* + \bP\, \btheta, \quad
  \btheta \in \mathbb R^D.
$$
Substituting into a second‑order Taylor expansion of $\cL_k$ around $\bw^*$ gives
$$
  \cL_k\bigl(\bw^* + \bP\,\btheta\bigr) \approx
  \cL_k(\bw^*) + \tfrac12\, \btheta^{\T}\! \bigl( \bP^{\T}\, \bH_k(\bw^*)\, \bP\bigr)\, \btheta =
  \cL_k(\bw^*) + \tfrac12\, \btheta^{\T} \bLambda_k\, \btheta,
$$
where
$$
  \bLambda_k =
  \bP^{\T}\, \bH_k(\bw^*)\, \bP =
  \operatorname{diag} \bigl( \lambda_k^{(1)}, \dots, \lambda_k^{(D)} \bigr).
$$
An identical expansion holds for $\cL_{k+1}$, yielding $\bLambda_{k+1}$ from the Hessian on $k+1$ samples.

\begin{theorem}[Approximate $\Delta_k$ via Eigenvalues]\label{th:th_1}
  Under the assumption that $\btheta \sim \mathcal N(0, \sigma^2 \mathbf I_D)$, the change in loss can be estimated by
  $$
    \Delta_k =
    \mathbb E \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2 \approx
    \frac{\sigma^4}{4}\, \Biggl( 2 \sum_{d=1}^D (\lambda_{k+1}^{(d)} - \lambda_k^{(d)})^2
    + \Bigl( \sum_{d=1}^D (\lambda_{k+1}^{(d)} - \lambda_k^{(d)}) \Bigr)^2 \Biggr).
  $$
\end{theorem}

@@

This formula uses only the top $D$ eigenvalues, so it is efficient to compute even for very large models.
In practice, it provides an upper bound on the empirical $\Delta_k$ and thus a reliable criterion for determining when the
dataset size is sufficient.

\section{Experiments}\label{sec:exp}

@@

\paragraph{Experimental Procedure.}
\begin{enumerate}
  \item \textbf{Random‐direction visualization \ref{sec:exp:pre}:} As an initial sanity check, we project the parameters onto two random directions and
        plot the resulting 3D loss landscapes, their squared differences, and the corresponding $\Delta_k$ for various settings.
  \item \textbf{Hessian‐based analysis \ref{sec:exp:hes}:} We repeat the above visualization using the top eigenvectors of the Hessian. This directly
        tests our hypothesis regarding the dominant role of the largest Hessian eigenvalues.
  \item \textbf{Main experiment \ref{sec:exp:main}:} We compute both the theoretical and empirical values of $\Delta_k$ and compare them. This evaluates
        the accuracy of our approximation predicted by theory.
\end{enumerate}

All code and detailed configuration files are available in our GitHub
repository \footnote{\href{https://github.com/intsystems/2025-Project-183}{GitHub}}.
\subsection{Preliminary Experiment: Projection onto Random Directions}\label{sec:exp:pre}
\subsubsection{Visualizing the Loss Landscape}

@@

First, a uniform grid of $(\alpha,\beta)$ pairs on the domain $[-1,1]\times[-1,1]$ with fixed step size $h$ is constructed.
For each grid point $(\alpha,\beta)$, we set $\mathbf{w} = \mathbf{w}(\alpha,\beta)$, keeping the base weights
$\mathbf{w}^*$ fixed, and compute the losses $\mathcal{L}_k(\mathbf{w})$ and $\mathcal{L}_{k+1}(\mathbf{w})$.
This produces two scalar fields over the grid: the first, $\mathcal{L}_k(\alpha,\beta)$, illustrates how the mean loss
varies across the parameter space, while the second,
$$\bigl(\mathcal{L}_{k+1}(\alpha,\beta) - \mathcal{L}_k(\alpha,\beta)\bigr)^2,$$ highlights the regions where adding a single
sample induces the largest change. Finally, both fields are visualized as two adjacent 3D plots.

@@

\paragraph{Loss Landscape for the Small Dataset}
In Figure \ref{fig:loss_random_small}, the loss surface is highly irregular, with sharp peaks and deep valleys.
This irregularity indicates that with very few training samples, the model is extremely sensitive to small parameter perturbations.
The squared difference surface also exhibits large spikes, confirming that adding a single data point can cause substantial local changes
in the loss.

\paragraph{Loss Landscape for the Full Dataset}
In contrast, Figure \ref{fig:loss_random_big} shows a much smoother loss surface with gentle slopes. The overall loss
values are lower --- reflecting the averaging effect of more samples --- and the squared difference surface is correspondingly much smaller.
This smoothing demonstrates that as the dataset grows, the local optimum stabilizes, and additional samples produce only minor
adjustments to the loss landscape. Hence, beyond a certain dataset size, further data have negligible impact on the shape of the loss
surface.

@@

To study how the loss landscape converges as the dataset grows, we estimate and plot the function
$$
  \Delta_k =
  \mathbb{E} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2
$$
via a Monte Carlo approximation using Gaussian perturbations in parameter space.

Starting from a fixed trained model to get $\mathbf{w}^*$, we draw $T$ independent random direction vectors
$$
  \btheta^{(t)} \sim \mathcal{N}(\mathbf{0}, \sigma^2 \mathbf{I}_d), \quad t = 1, \dots, T.
$$
For each batch, we form a perturbed parameter set $\mathbf{w}^{(t)} = \mathbf{w}^* + \btheta^{(t)}$,
and evaluate the losses $\mathcal{L}_k(\mathbf{w}^{(t)})$ for each $k = 1, \, \dots, \, N$,
after what compute the squared difference on given values:
$$
  d_t = \bigl[\mathcal{L}_{k+1}(\mathbf{w}^{(t)}) - \mathcal{L}_k(\mathbf{w}^{(t)})\bigr]^2.
$$
We then approximate the average change
$\Delta_k \approx \frac{1}{T} \sum_{t=1}^T d_t$,
noting that increasing $T$ reduces the variance of this Monte Carlo estimator. Finally, we plot $\Delta_k$ against $k$ to observe its decay
and include the rescaled quantity $\Delta_k \cdot k^2$ to highlight the expected $\mathcal{O}(1/k)$ convergence rate.

Figures \ref{fig:delta_random_sigma} and \ref{fig:delta_random_dim} present the behavior of the function
$$
  \Delta_k =
  \mathbb{E} \bigl[ \cL_{k+1}(\bw) - \cL_k(\bw) \bigr]^2
$$
under different Monte Carlo sampling settings and projection subspace sizes. Each figure isolates one variable:

\begin{itemize}
  \item Figure \ref{fig:delta_random_sigma}: varying the Gaussian variance $\sigma$.
  \item Figure \ref{fig:delta_random_dim}: varying the subspace dimension $D$.
\end{itemize}

\begin{figure}[!htbp]
  \hspace*{-2.6cm}
  \subfloat{\includegraphics[width=1.35\textwidth]{img/delta_random_sigma_2_1024.pdf}}
  \caption{$\Delta_k$ (left) and $\Delta_k \cdot k^2$ (right), varying the Gaussian variance with fixed $d = 2$.}
  \label{fig:delta_random_sigma}
\end{figure}

\begin{figure}[!htbp]
  \hspace*{-2.6cm}
  \subfloat{\includegraphics[width=1.35\textwidth]{img/delta_random_dim_1_1024.pdf}}
  \caption{$\Delta_k$ (left) and $\Delta_k \cdot k^2$ (right), varying the subspace dimension with fixed $\sigma = 1$.}
  \label{fig:delta_random_dim}
\end{figure}

\paragraph{Comparing Different Variances \texorpdfstring{$\sigma$}{sigma}}
Figure \ref{fig:delta_random_sigma} shows that a high Gaussian variance $\sigma$ produces a significant change in both the shape
and magnitude of the $\Delta_k$ curves. This occurs because, as we move away from the local minimum, the value of the loss
function changes. However, this change is modest: the insensitivity likely stems from using random projection directions,
which do not align with the dominant curvature axes of the Hessian, so scaling those directions has little impact on the estimated
loss differences. Moreover, variances not exceeding one yield very similar effects, so we choose $\sigma = 1$ for simplicity in our
calculations.

\paragraph{Comparing Different Subspace Dimensions}
Figure \ref{fig:delta_random_dim} shows that, when using random directions, increasing the projection dimension $D$ does not
significantly alter the overall shape of $\Delta_k$. This suggests that only a few directions dominate the variation --- motivating our
focus on Hessian‐based projections in the main experiment.

In all cases, the right‐hand plots (showing $\Delta_k \cdot k^2$) exhibit a roughly $\mathcal O(1 / k^2)$ decay, confirming that the
incremental effect of adding one more sample diminishes proportionally to $1 / k$ as the dataset grows.

@@

\subsection{Moving Forward: Projection onto Dominant Hessian Eigenvectors}\label{sec:exp:hes}

\paragraph{Projection Directions.}
In the main experiment, we replace random directions with the $D$ eigenvectors of the Hessian at the local minimizer $\bw^*$ that
correspond to its largest eigenvalues. Prior work \cite{sagun2018empirical} shows these top modes capture nearly all of the
non‑negligible curvature, and Figure \ref{fig:evgen} confirms a rapid eigenvalue decay. Fixing $D = 10$ balances computational cost and
fidelity, and our tests on Fashion‑MNIST demonstrate that this choice suffices even for complex architectures.

@@

Figures \ref{fig:loss_eigen_small} and \ref{fig:loss_eigen_big} display 3D plots of $\cL_k(\btheta)$ (left) and
$\bigl[ \cL_{k+1}(\btheta) - \cL_k(\btheta) \bigr]^2$ (right), where $\btheta$ varies over the span of the top two Hessian eigenvectors.
The procedure mirrors the random‑direction case but uses these maximally informative axes.

@@

\paragraph{Loss Landscape for the Small Dataset}
In Figure \ref{fig:loss_eigen_small}, the loss surface projected onto the top two Hessian eigenvectors is
noticeably more structured and less erratic than in the random‐direction case. This confirms that these eigenvectors capture the
most influential directions of loss variation. The squared‐difference plot also shows larger peaks, indicating that the model is
highly sensitive to new data along these key axes when $k$ is small.

\paragraph{Loss Landscape for the Full Dataset}
By contrast, Figure \ref{fig:loss_eigen_big} shows a much smoother loss surface and a greatly reduced squared‐difference.
As the dataset grows, the loss changes less along the principal curvature directions, and the model’s local behavior stabilizes.
Notably, even at large $k$, the squared difference remains higher than in the random‐direction experiment, providing a practical upper
bound on landscape fluctuations. These findings reinforce that projecting onto the top eigenvectors effectively reveals the convergence
behavior of the loss landscape and helps intuitively understand the existence of the point beyond which additional data have minimal impact.

@@

\begin{figure}[!htbp]
  \hspace*{-2.6cm}
  \subfloat{\includegraphics[width=1.35\textwidth]{img/delta_eigen_1_10_1024.pdf}}
  \caption{$\Delta_k$ (left) and $\Delta_k \cdot k^2$ (right)}
  \label{fig:delta_eigen}
\end{figure}

Figure \ref{fig:delta_eigen} presents the Monte Carlo estimate of $\Delta_k$ in the subspace spanned by the top $d = 10$
Hessian eigenvectors. Compared to the random‐direction experiment, $\Delta_k$ here has a higher value --- reflecting the
large curvature captured by these eigenvectors. The right‐hand plot of $\Delta_k\cdot k^2$
confirms an approximate $\mathcal O(1 / k^2)$ convergence rate. These observations show that projecting onto the leading eigen-directions not only
highlights the greatest loss sensitivity but also yields a reliable upper bound on how quickly the landscape stabilizes as more data are
added.

\subsection{Main Experiment: Theoretical vs. Empirical \texorpdfstring{$\Delta_k$}{Delta k}}\label{sec:exp:main}

@@

As the Theorem \ref{th:th_1} says, we approximate $\Delta_k$ by
$$
  \Delta_k \approx
  \frac{\sigma^4}{4}\, \Biggl( 2 \sum_{d=1}^D (\lambda_{k+1}^{(d)} - \lambda_k^{(d)})^2
  + \Bigl( \sum_{i=1}^D (\lambda_{k+1}^{(d)} - \lambda_k^{(d)}) \Bigr)^2 \Biggr).
$$
where $\{\lambda_k^{(i)}\}_{i=1}^d$ and $\{\lambda_{k+1}^{(i)}\}_{i=1}^d$ are the top $D$ Hessian eigenvalues at sample sizes $k$ and $k + 1$,
and $\sigma$ is the perturbation variance.

We compute both empirical and theoretical estimates of $\Delta_k$. First, we train a model instance as described above.
For each batch $k$, we apply the power‐iteration method to the model to extract the top $D$ Hessian eigenvalues
$\{\lambda_k^{(d)}\}_{d=1}^D$. The theoretical estimate of $\Delta_k$ is then obtained by substituting these eigenvalues into our
approximation. To reduce high-frequency fluctuations, we apply a five‐point moving average to the sequence:
$\frac{1}{5}\bigl(\Delta_{k-2} + \Delta_{k-1} + \Delta_k + \Delta_{k+1} + \Delta_{k+2}\bigr)$ for each $k$,
where, at the boundaries, any index outside $[1,N]$ is replaced by the nearest valid endpoint (i.e.\ we replicate the border value),
which is equivalent to convolution with constant‐value padding.

In parallel, we compute empirical estimates of $\Delta_k$ via Monte Carlo sampling of Gaussian perturbations along the same
top-$D$ eigenvectors, following the procedure described above. Finally, we plot the
theoretical and empirical $\Delta_k$ on the same axes, allowing us to assess how closely the eigenvalue-based bound tracks the observed
convergence of the loss landscape.

\begin{figure}[!htbp]
  \subfloat{\includegraphics[width=1\textwidth]{img/delta_border_1_10_1024.pdf}}
  \caption{Empirical and Theoretical $\Delta_k$}
  \label{fig:delta_eigen_comparison}
\end{figure}

Figure \ref{fig:delta_eigen_comparison} shows the theoretical estimate of $\Delta_k$ (dashed line) alongside the empirical Monte Carlo
measurements (solid line). Across all values of $k$, the theory curve remains above the empirical one, demonstrating that our
eigenvalue‐based formula provides a valid upper bound for the true loss‐landscape changes.

Now, to eliminate the explicit batch index $k$, we plot $\Delta_k$ directly against $\cL_k$ and apply the same moving average
for a smoother, more stable curve.

Figure \ref{fig:delta_loss} shows the relationship between the loss $\cL_k(\bw^*)$ and the corresponding change $\Delta_k(\bw^*)$.
In the top panel, each blue dot represents the empirical estimate of $\Delta_k$ versus $\cL_k$, and the red line is the best-fit
linear trend. The bottom panel shows the analogous theoretical estimates, again smoothed and fitted (green line). In both cases,
there is a strong positive correlation: higher loss values correspond to larger observed and predicted changes. Although some
scatter remains due to sampling noise, the overall linear relationship validates our hypothesis that the eigenvalue-based
bound $\Delta$ captures the dominant curvature effects of the Hessian. Consequently, the $\Delta$-sufficient condition can effectively
limit changes in the loss function.

\begin{figure}[!htbp]
  \subfloat{\includegraphics[width=1\textwidth]{img/delta_loss_1_10_1024.pdf}}
  \caption{Empirical and Theoretical $\Delta_k$}
  \label{fig:delta_loss}
\end{figure}
